Statistical Estimation of Mixing Times

### 2015-06-01 scratch/old-notes.tex
Copied files from svn

\newcommand\wh{\ensuremath{\widehat}}
\newcommand\wt{\ensuremath{\widetilde}}
\newcommand\norm[1]{\| #1 \|}
\newcommand\Norm[1]{\left\| #1 \right\|}
\newcommand\tvnorm[1]{\left\| #1 \right\|_{\mathrm{TV}}}
\newcommand\Parens[1]{\left( #1 \right)}
\newcommand\parens[1]{( #1 )}
\newcommand\Abs[1]{\left| #1 \right|}
\newcommand\abs[1]{| #1 |}
\DeclareMathOperator{\Diag}{Diag}
\DeclareMathOperator{\Sym}{Sym}
\renewcommand\t{{\ensuremath{\scriptscriptstyle{\top}}}}
\newcommand\be{\ensuremath{\mathbf{e}}}
\newcommand\tmix{\ensuremath{t_{\mathrm{mix}}}}
\newcommand\htmix{\ensuremath{\hat{t}_{\mathrm{mix}}}}
\newcommand{\od}{\bar{d}}
\newcommand{\tcouple}{\tau_{\mathrm{couple}}}
\newcommand{\ZZ}{\mathcal{Z}}
\newcommand\trel{\ensuremath{t_{\mathrm{rel}}}}
\newcommand{\ip}[1]{\langle#1\rangle}
\newcommand{\DF}{\mathcal{E}}
\newcommand\ttmix{\ensuremath{t_{\mathrm{mix},2}}}

@@

Define the $\pi$-weighted inner product by $\ip{u,v}_{\pi} = \sum_i u_i v_i \pi_i$ and denote the corresponding norm by $\norm{\cdot}_{2,\pi}$. More generally, we denote the $\pi$-weighted $p$-norm by $\norm{\cdot}_{p,\pi}$.
Let $T$ be a reversible transition matrix, with eigenvalues $1 = \lambda_1 > \lambda_2 \ge \ldots \lambda_d \ge -1$ and associated eigenvectors $\{v_j\}_j$ with respect to $\ip{\cdot,\cdot}_\pi$. Then,
\[
\tvnorm{ T^t(i,\cdot) - \pi}^2 
= \frac14  \Norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{1,\pi}^2  
\le \frac14  \Norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}^2
= \frac14 \sum_{j=2}^d v_j^2 \lambda_j^{2t}\,
\]
(cf. Lemma~12.16 of \citet{LePeWi08}).
\if0
Furthermore, if the chain is transitive then 
\[
\tvnorm{ T^t(i,\cdot) - \pi}^2 \le \frac14  \Norm{\frac{T^t(i,\cdot)}{\pi_i} - \be }_2^2 \frac14 \sum_{j=2}^d \lambda_j^{2t}\,.
\]
\fi
We shall let $d_2(t) =  \max_i \Norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}$ and the $\ell^2$ mixing time by $\ttmix(\eps) = \min\{ t: d_2(t)\le \eps\}$.
Denote $\norm{T^*} = \sup_{v\ne 0, v^\top \pi = 0} \frac{\norm{T^* v}_{2,\pi}}{\norm{v}_{2,\pi}}$.
Then, 
\[
\ttmix(\eps) \le \ceil{ \frac{1}{1-\norm{T^*}} \ln \frac{1}{\eps\sqrt{\pi_*}}} 
%=  \ceil{\frac1{\gamma_{T T^*}}\ln \frac1{\eps\sqrt{\pi_*}}} \,,
\]
(cf. Proposition~1.2 \citet{MoTe06}). 
\if0
Here, for a Markov chain with kernel $K$ and stationary distribution $\pi$,
 $\gamma_{K}$ is defined by
\[
\gamma_{K} = \inf_v \frac{\mathrm{Var}_\pi(v)}{\DF_K(v,v)}\,,
\]
where $\mathrm{Var}_\pi(v) = \sum_i (v_i - \ip{v,\pi})^2 \pi_i$ and
$\DF_K(u,v) = \ip{ u, (I-K)v }_\pi$ is the \emph{Dirichlet form} underlying $K$.
If $T$ is reversible, $\gamma_{T T^*} =\gamma_{T^2} = \gamma_*^2$.
\fi
If the chain is non-reversible, the above bound may be trivial (even though the chain may be contracting). For example, this happens when $d = 3$, $T_{1,2} = T_{2,3} = 1$, $T_{3,3} = T_{3,1} = 1/2$. The origin of the failure of the bound is that the methods require that the distance decreases at every step, while if the initial distribution is concentrated at state $1$, then it will take two steps before the distance decreases. This is a common problem with the bounds of this type for the nonreversible case.
%Then, $\pi_1 = \pi_2 = 1/4$, $\pi_3 = 1/2$.

\section{The matrix estimation method}

@@

Let $P := \Diag(\pi) T$ (i.e., $P_{i,j} := \pi_i T_{i,j}$), so $L =
\Diag(\pi)^{-1/2} P \Diag(\pi)^{-1/2}$.
We are interested in approximating the eigenvalues of the matrix $L :=
\Diag(\pi)^{1/2} T \Diag(\pi)^{-1/2}$ (which is symmetric by
reversibility).

Our approach is to directly estimate $P$ and $\pi$ to form an estimate
$\wh{L}$ of $L$.
Then, we use the eigenvalues of $\wh{L}$ to form our mixing time
estimate.

@@

Let $\wh{P}$ and $\hat\pi$ be estimates of $P$ and $\pi$,
respectively, and define
\begin{align*}
  \wh{L} & := \Diag(\hat\pi)^{-1/2} \wh{P} \Diag(\hat\pi)^{-1/2} \\
  E_\pi & := \Diag(\hat\pi)^{-1/2} \Diag(\pi)^{1/2} - I \\
  E_P & := \Diag(\pi)^{-1/2} (\wh{P} - P) \Diag(\pi)^{-1/2} .
\end{align*}
Then
\[
  \wh{L} - L
  = E_P + E_\pi L + L E_\pi + E_\pi L E_\pi + E_\pi E_P + E_P E_\pi +
  E_\pi E_P E_\pi
  ,
\]
so
\[
  \norm{\wh{L} - L}_2
  \leq \norm{E_P}_2 + (\norm{E_P}_2 + \norm{L}_2)
  (2\norm{E_\pi}_2 + \norm{E_\pi}_2^2)
  .
\]

Recall that $L$ is symmetric and that its eigenvalues are conained in
$[-1,1]$.
This is because $I - L$ is a normalized Laplacian matrix, which has
eigenvalues contained in the interval $[0,2]$.
In fact, the smallest eigenvalue of $I - L$ is zero, so the largest
eigenvalue of $L$ is equal to $1$.
Therefore $\norm{L}_2 = \max\{\lambda_{\max}(L),-\lambda_{\min}(L)\} =
1$, and
\[
  \norm{\wh{L} - L}_2
  \leq \norm{E_P}_2 + (\norm{E_P}_2 + 1)
  (2\norm{E_\pi}_2 + \norm{E_\pi}_2^2)
  .
\]

Let $\Sym(A) := 0.5 (A + A^\t)$ for a square matrix $A$.
By the triangle inequality,
\[
  \norm{\Sym(\wh{L}) - L}_2
  \leq \norm{\wh{L} - L}_2 .
\]
Let $\lambda_i(A)$ denote the $i$-th eigenvalue of a symmetric matrix
$A$ (assuming the eigenvalues are arranged in non-increasing order).
By Weyl's inequality,
\[
  |\lambda_i(\Sym(\wh{L})) - \lambda_i(L)| \leq
  \norm{\wh{L} - L}_2
\]
for all $i\in[d]$.

@@

Suppose we have $n$ iid copies of $(X_t,X_{t+1})$, where the marginal
distribution of $X_t$ is $\pi$, and $\wh{P}$ is the maximum likelihood
estimate of $P$.
Then we can write $E_P$ as the average of $n$ independent copies of the
zero-mean random matrix $Z$, where
\[
  \Pr\Parens{ Z = \frac1{\sqrt{\pi_i\pi_j}} e_i e_j^\t - L } = P_{i,j} .
\]
We can apply the matrix Bernstein inequality to bound $\norm{E_P}_2$ with
high probability.
To do this, we need to bound the following quantities:
\[
  \norm{Z}_2 , \quad
  \norm{\bbE(ZZ^\t)}_2 , \quad
  \norm{\bbE(Z^\t Z)}_2 .
\]
By the triangle inequality, we have
\[
  \norm{Z}_2
  \leq \frac{1}{\pi_{\min}} + \norm{L}_2
  = \frac{1}{\pi_{\min}} + 1
\]
where $\pi_{\min} := \min_{i\in[d]} \pi_i$.
Moreover, by reversibility,
\[
  \bbE(ZZ^\t)
  = \bbE(Z^\t Z)
  = \sum_{i=1}^d \Parens{ \sum_{j=1}^d \frac{T_{i,j}}{\pi_j} }
  e_ie_i^\t - L^2
\]
so
\[
  \norm{\bbE(ZZ^\t)}_2
  = \norm{\bbE(Z^\t Z)}_2
  \leq \max_{i \in [d]} \sum_{j=1}^d \frac{T_{i,j}}{\pi_j} + \norm{L}_2^2
  = D_{T,\pi} + 1
\]
where $D_{T,\pi} := \max_{i\in[d]} \sum_{j=1}^d T_{i,j} / \pi_j \leq
1/\pi_{\min}$.
By the matrix Bernstein inequality, there is an absolute constant
$c>0$ (maybe $2$ or so) such that with probability at least
$1-\delta$,
\begin{align*}
  \norm{E_P}_2
  & \leq c \sqrt{\Parens{ \frac1{\pi_{\min}} + 1 }\frac{\ln(4d/\delta)}{n}}
  + c \Parens{ \frac1{\pi_{\min}} + 1 } \frac{\ln(4d/\delta)}{n}
  .
\end{align*}

We can also estimate $\pi$ using a maximum likelihood estimator
$\wh\pi$.
We want a bound on
\[
  \norm{E_\pi}_2 = \max_{i\in[d]} \Abs{\sqrt{\pi_i/\wh\pi_i} - 1}
  .
\]
If we are willing to assume that $n \geq c / \pi_{\min}$ for some
suitable constant $c>0$, then by multiplicative Chernoff bounds and a
union bound, we have with probability at least $1-\delta$,
\[
  \norm{E_\pi}_2 \leq
  c' \sqrt{\frac{\ln(d/\delta)}{\pi_{\min} n}} +
  + c' \frac{\ln(d/\delta)}{\pi_{\min} n}
  .
\]
We might be able to avoid this assumption by using something like
``inverse sampling''.

\bibliographystyle{plainnat}
\bibliography{../all}

>>>>

\newcommand\wh{\ensuremath{\widehat}}
\newcommand\wt{\ensuremath{\widetilde}}
\newcommand\norm[1]{\left\| #1 \right\|}
\newcommand\tvnorm[1]{\left\| #1 \right\|_{\mathrm{TV}}}
\newcommand\Parens[1]{\left( #1 \right)}
\DeclareMathOperator{\Diag}{Diag}
\DeclareMathOperator{\Sym}{Sym}
\renewcommand\t{{\ensuremath{\scriptscriptstyle{\top}}}}
\newcommand\be{\ensuremath{\mathbf{e}}}
\newcommand\tmix{\ensuremath{t_{\mathrm{mix}}}}
\newcommand\htmix{\ensuremath{\hat{t}_{\mathrm{mix}}}}
\newcommand{\od}{\bar{d}}
\newcommand{\tcouple}{\tau_{\mathrm{couple}}}
\newcommand{\ZZ}{\mathcal{Z}}
\newcommand\trel{\ensuremath{t_{\mathrm{rel}}}}
\newcommand{\ip}[1]{\langle#1\rangle}
\newcommand{\DF}{\mathcal{E}}
\newcommand\ttmix{\ensuremath{t_{\mathrm{mix},2}}}

@@

Define the $\pi$-weighted inner product by $\ip{u,v}_{\pi} = \sum_i u_i v_i \pi_i$ and denote the corresponding norm by $\norm{\cdot}_{2,\pi}$. More generally, we denote the $\pi$-weighted $p$-norm by $\norm{\cdot}_{p,\pi}$.
Let $T$ be a reversible transition matrix, with eigenvalues $1 = \lambda_1 > \lambda_2 \ge \ldots \lambda_d \ge -1$ and associated eigenvectors $\{v_j\}_j$ with respect to $\ip{\cdot,\cdot}_\pi$. Then,
\[
\tvnorm{ T^t(i,\cdot) - \pi}^2 
= \frac14  \norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{1,\pi}^2  
\le \frac14  \norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}^2
= \frac14 \sum_{j=2}^d v_j^2 \lambda_j^{2t}\,
\]
(cf. Lemma~12.16 of \citet{LePeWi08}).
\if0
Furthermore, if the chain is transitive then 
\[
\tvnorm{ T^t(i,\cdot) - \pi}^2 \le \frac14  \norm{\frac{T^t(i,\cdot)}{\pi_i} - \be }_2^2 \frac14 \sum_{j=2}^d \lambda_j^{2t}\,.
\]
\fi
We shall let $d_2(t) =  \max_i \norm{\frac{T^t(i,\cdot)}{\pi(\cdot)} - \be }_{2,\pi}$ and the $\ell^2$ mixing time by $\ttmix(\eps) = \min\{ t: d_2(t)\le \eps\}$.
Denote $\norm{T^*} = \sup_{v\ne 0, v^\top \pi = 0} \frac{\norm{T^* v}_{2,\pi}}{\norm{v}_{2,\pi}}$.
Then, 
\[
\ttmix(\eps) \le \ceil{ \frac{1}{1-\norm{T^*}} \ln \frac{1}{\eps\sqrt{\pi_*}}} 
%=  \ceil{\frac1{\gamma_{T T^*}}\ln \frac1{\eps\sqrt{\pi_*}}} \,,
\]
(cf. Proposition~1.2 \citet{MoTe06}). 
\if0
Here, for a Markov chain with kernel $K$ and stationary distribution $\pi$,
 $\gamma_{K}$ is defined by
\[
\gamma_{K} = \inf_v \frac{\mathrm{Var}_\pi(v)}{\DF_K(v,v)}\,,
\]
where $\mathrm{Var}_\pi(v) = \sum_i (v_i - \ip{v,\pi})^2 \pi_i$ and
$\DF_K(u,v) = \ip{ u, (I-K)v }_\pi$ is the \emph{Dirichlet form} underlying $K$.
If $T$ is reversible, $\gamma_{T T^*} =\gamma_{T^2} = \gamma_*^2$.
\fi
If the chain is non-reversible, the above bound may be trivial (even though the chain may be contracting). For example, this happens when $d = 3$, $T_{1,2} = T_{2,3} = 1$, $T_{3,3} = T_{3,1} = 1/2$. The origin of the failure of the bound is that the methods require that the distance decreases at every step, while if the initial distribution is concentrated at state $1$, then it will take two steps before the distance decreases. This is a common problem with the bounds of this type for the nonreversible case.
%Then, $\pi_1 = \pi_2 = 1/4$, $\pi_3 = 1/2$.

\section{Analysis via Estimation of $T$}

@@

\subsection{Estimation error}

We use the median-of-means technique to estimate all $d^2$ entries of
$T$ (this is just to get a variance-only bound with high probability;
using the empirical average also works but will involve some
additional lower-order terms).
That is, for each $(i,j)\in[d]^2$, we consider the $\hat{n}_i$
times $t$ where $X_t = i$ and split them into $k$ groups each of size
$\hat{n}_i/k$; within each group, we compute the fraction of these
times where $X_{t+1}=j$; we then compute the median of these fractions
and use that as the estimate of $T_{i,j}$.
Call these estimates $\wt{T}_{i,j}$.
Using $k := 4.5\ln(d^2/\delta)$, we have using the median-of-means
theorem and a union bound,
\[
  \Parens{ \wt{T}_{i,j} - T_{i,j} }^2
  \leq \frac{27T_{i,j}(1-T_{i,j})\ln(d^2/\delta)}{\hat{n}_i}
\]
for all $(i,j) \in [d]^2$, with probability $\geq1-\delta$.
Therefore, in this event,
\begin{align*}
  \norm{\Diag(\pi)^{1/2} (\wt{T} - T) \Diag(\pi)^{-1/2}}_F^2
  & = \sum_{i=1}^d \sum_{j=1}^d \frac{\pi_i}{\pi_j}
  \Parens{ \wt{T}_{i,j} - T_{i,j} }^2 \\
  & \leq \sum_{i=1}^d \sum_{j=1}^d
  \frac{27\pi_i T_{i,j} (1 - T_{i,j})\ln(d^2/\delta)}{\pi_j\hat{n}_i} \\
  & = \sum_{i=1}^d \sum_{j=1}^d
  \frac{27\pi_j T_{j,i} (1 - T_{i,j})\ln(d^2/\delta)}{\pi_j\hat{n}_i} \\
  & = \sum_{i=1}^d \sum_{j=1}^d
  \frac{27T_{j,i} (1 - T_{i,j})\ln(d^2/\delta)}{\hat{n}_i} \\
  & = \sum_{i=1}^d \frac{27d_i\ln(d^2/\delta)}{\hat{n}_i}
\end{align*}
where $d_i := \sum_{j=1}^d T_{j,i} (1 - T_{i,j})$.

@@

We are interested in approximating the eigenvalues of the matrix $L :=
\Diag(\pi)^{1/2} T \Diag(\pi)^{-1/2}$ (which is symmetric by
reversibility).

Let $\wt{T}$ and $\tilde\pi$ be estimates of $T$ and $\pi$,
respectively.
%Assume the Markov chain associated with $\wt{T}$ has a stationary
%distribution denoted by $\tilde\pi$.
%Since $I - T + 1\pi^\t$ is invertible and
%\[
%  (\tilde\pi - \pi)^\t (I - T + 1\pi^\t)
%  = \tilde\pi^\t (\wt{T} - T)
%  ,
%\]
%we can bound $\|\tilde\pi - \pi\|_\infty$ by
%\[
%  \|\tilde\pi - \pi\|_\infty
%  \leq \|(\wt{T} - T)(I - T + 1\pi^\t)^{-1}\|_{\infty\to1}
%  .
%\]
Define
\begin{align*}
  \wt{L}
  & := \Diag(\tilde\pi)^{1/2} \wt{T} \Diag(\tilde\pi)^{-1/2}
  \\
  E_\pi & := \Diag(\pi)^{1/2} \Diag(\tilde\pi)^{-1/2} - I \\
  E_\pi' & := \Diag(\tilde\pi)^{1/2} \Diag(\pi)^{-1/2} - I \\
  E_T & := \Diag(\pi)^{1/2} (\wt{T} - T) \Diag(\pi)^{-1/2} .
\end{align*}
Then
\[
  \wt{L} - L
  = E_T + E_\pi' L + L E_\pi + E_\pi' L E_\pi
  + E_\pi' E_T + E_T E_\pi + E_\pi' E_T E_\pi
  ,
\]
so
\begin{align*}
  \|\wt{L} - L\|_2
  & \leq \|E_T\|_2
  + (\|L\|_2 + \|E_T\|_2)
  (\|E_\pi\|_2 + \|E_\pi'\|_2 + \|E_\pi\|_2\|E_\pi'\|_2)
  .
\end{align*}

Let $\Sym(A) := 0.5 (A + A^\t)$ for a square matrix $A$.
By the triangle inequality,
\[
  \norm{\Sym(\wt{L}) - L}_2
  \leq \|\wt{L} - L\|_2 .
\]
Let $\lambda_i(A)$ denote the $i$-th eigenvalue of a symmetric matrix
$A$ (assuming the eigenvalues are arranged in non-increasing order).
By Weyl's inequality,
\[
  |\lambda_i(\Sym(\wt{L})) - \lambda_i(L)| \leq
  \|\wt{L} - L\|_2
\]
for all $i\in[d]$.

\section{Analysis via Estimation of $\Diag(\pi)T$}

Let $P := \Diag(\pi) T$ (i.e., $P_{i,j} := \pi_i T_{i,j}$), so $L =
\Diag(\pi)^{-1/2} P \Diag(\pi)^{-1/2}$.

Let $\wh{P}$ and $\hat\pi$ be estimates of $P$ and $\pi$,
respectively, and define
\begin{align*}
  \wh{L} & := \Diag(\hat\pi)^{-1/2} \wh{P} \Diag(\hat\pi)^{-1/2} \\
  E_\pi & := \Diag(\hat\pi)^{-1/2} \Diag(\pi)^{1/2} - I \\
  E_P & := \Diag(\pi)^{-1/2} (\wh{P} - P) \Diag(\pi)^{-1/2} .
\end{align*}
Then
\[
  \wh{L} - L
  = E_P + E_\pi L + L E_\pi + E_\pi L E_\pi + E_\pi E_P + E_P E_\pi +
  E_\pi E_P E_\pi
  ,
\]
so
\[
  \|\wh{L} - L\|_2
  \leq \|E_P\|_2 + (\|E_P\|_2 + \|L\|_2) (2\|E_\pi\|_2 + \|E_\pi\|_2^2)
  .
\]

Recall that $L$ is symmetric and that its eigenvalues are between $-1$
and $+1$.
This is because $I - L$ is a normalized Laplacian matrix, which has
eigenvalues contained in the interval $[0,2]$.
In fact, the smallest eigenvalue of $I - L$ is zero, so the largest
eigenvalue of $L$ is equal to $1$; the smallest eigenvalue of $L$ is
at least $-1$.
Therefore $\|L\|_2 = \max\{\lambda_{\max}(L),-\lambda_{\min}(L)\} =
1$.
Therefore
\[
  \|\wh{L} - L\|_2
  \leq \|E_P\|_2 + (\|E_P\|_2 + 1) (2\|E_\pi\|_2 + \|E_\pi\|_2^2)
  .
\]

@@

Suppose we have $n$ iid copies of $(X_t,X_{t+1})$, where the marginal
distribution of $X_t$ is $\pi$, and $\wh{P}$ is the maximum likelihood
estimate of $P$.
Then we can write $E_P$ as the average of $n$ independent copies of the
zero-mean random matrix $Z$, where
\[
  \Pr\Parens{ Z = \frac1{\sqrt{\pi_i\pi_j}} e_i e_j^\t - L } = P_{i,j} .
\]
We can apply the matrix Bernstein inequality to bound $\|E_P\|_2$ with
high probability.
To do this, we need to bound the following quantities:
\[
  \|Z\|_2 , \quad
  \|\bbE(ZZ^\t)\|_2 , \quad
  \|\bbE(Z^\t Z)\|_2 .
\]
By the triangle inequality, we have
\[
  \|Z\|_2
  \leq \frac{1}{\pi_{\min}} + \|L\|_2
  = \frac{1}{\pi_{\min}} + 1
\]
where $\pi_{\min} := \min_{i\in[d]} \pi_i$.
Moreover, by reversibility,
\[
  \bbE(ZZ^\t)
  = \bbE(Z^\t Z)
  = \sum_{i=1}^d \Parens{ \sum_{j=1}^d \frac{T_{i,j}}{\pi_j} }
  e_ie_i^\t - L^2
\]
so
\[
  \|\bbE(ZZ^\t)\|_2
  = \|\bbE(Z^\t Z)\|_2
  \leq \max_{i \in [d]} \sum_{j=1}^d \frac{T_{i,j}}{\pi_j} + \|L\|_2^2
  = D_{T,\pi} + 1
\]
where $D_{T,\pi} := \max_{i\in[d]} \sum_{j=1}^d T_{i,j} / \pi_j \leq
1/\pi_{\min}$.
By the matrix Bernstein inequality, there is an absolute constant
$c>0$ such that with probability at least $1-\delta$,
\begin{align*}
  \|E_P\|_2
  & \leq c \sqrt{\frac{(D_{T,\pi} + 1) \ln(d/\delta)}{n}}
  + c \Parens{ \frac1{\pi_{\min}} + 1 } \frac{\ln(d/\delta)}{n}
  \\
  & \leq c \sqrt{\Parens{ \frac1{\pi_{\min}} + 1 }\frac{\ln(d/\delta)}{n}}
  + c \Parens{ \frac1{\pi_{\min}} + 1 } \frac{\ln(d/\delta)}{n}
  .
\end{align*}

\bibliographystyle{plainnat}
\bibliography{all}
